---
title: "Final Project - Week 6"
author: "Shrey Kothari"
date: "2025-11-04"
output: html_document
---

Step 1:
```{r}
library(tidyverse)
data <- read_csv("cox-violent-parsed.csv")
glimpse(data)
```
```{r}
set.seed(862331611)
rows <- nrow(data)
trainIndex <- sample(1:rows, size = 0.7*rows)
trainData <- data[trainIndex, ]
testData <- data[-trainIndex, ]
trainData <- trainData %>%
  rename(priors_count = priors_count...15)
testData <- testData %>%
  rename(priors_count = priors_count...15)
```

Step 2:

The csv file is the COMPAS dataset that looks at the probability of a criminial commiting another crime in the future. It comes from https://www.kaggle.com/datasets/danofer/compass/data. After a 70/30 split using seed = 862331611, the training set contains 12,821 observations and the test set contains 5,495.
We have 52 variables in total.
The response variable is "is_recid" while the candidate predictors are race, sex, age, priors_count, c_charge_degree, decile_score, and juv_fel_count. 

The is_recid variable is boolean and tells us whether a person will commit a crime in the future. 
The race variable is type string, which tells us the race of the person.
The sex variable is type string, which tells us the person's sex.
The priors_count variable is type int, which tells us how many prior counts the person has.
The juv_fel_count variable is type int, which tells us how many felonys a person had as a juvenile.
The decile_score variable is type int, which tells us perceived risk of recidivism.
The c_charge_degree is type string, which tells us the charge degree of a person's crime.

Step 3:
```{r}
variablesOfInterest <- c("is_recid", "race", "age", "priors_count")
df <- trainData[, variablesOfInterest]

numericVariables <- c("age", "priors_count")
categoricalVariables <- c("race", "is_recid")
```
```{r}
numericSummary <- data.frame(
  Variable = numericVariables,
  Mean = sapply(df[numericVariables], mean, na.rm = TRUE),
  Median = sapply(df[numericVariables], median, na.rm = TRUE),
  SD = sapply(df[numericVariables], sd, na.rm = TRUE),
  Min = sapply(df[numericVariables], min, na.rm = TRUE),
  Q1 = sapply(df[numericVariables], quantile, probs = 0.25, na.rm = TRUE),
  Q3 = sapply(df[numericVariables], quantile, probs = 0.75, na.rm = TRUE),
  Max = sapply(df[numericVariables], max, na.rm = TRUE)
)
numericSummary
```
```{r}
for (v in categoricalVariables) {
  cat("\nVariable:", v, "\n")
  tbl <- table(df[[v]])
  pct <- prop.table(tbl) * 100
  print(cbind(Frequency = tbl,
              Percentage = round(pct, 1)))
}
```
From the information, age seems to be right-skewed. The median is 31 years, while the max is 96 years. For prior_counts, the Q1 is equal to 0. This tells me that a lot of people have 0 prior counts. For race, we have a large number of African-Americans and Caucasians. 


Step 4:

```{r}
isRecidPlot <- barplot(table(df$is_recid),
                      main = "Recid Distribution",
                      xlab = "Recidivism",
                      ylab = "Count",
                      col = c("blue", "red", "green"))

racePlot <- barplot(table(df$race),
                    main = "Race Distribution",
                    xlab = "Race",
                    ylab = "Count")
```
```{r}
ageHistogram <- hist(table(df$age),
                      main = "Age Distribution",
                      xlab = "Age",
                      ylab = "Count",
                      col = "purple")

priorCountsHistogram <- hist(table(df$priors_count),
                            main = "Prior Counts Distribution",
                            xlab = "Prior Counts",
                            ylab = "Count",
                            col = "orange")
```
We can see in the prior_counts histogram that we have a large count where the number of prior counts is less than or equal to 1000. We do observe some outliers in the prior_counts, where we have less than 3 people who have committed around 3500-4000 prior counts. 

```{r}
library(ggplot2)

ggplot(df, aes(x = age, y = as.numeric(is_recid))) +
  geom_jitter(height = 0.05, alpha = 0.3) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  theme_bw() +
  labs(title = "is_recid vs Age", y = "is_recid")
```
```{r}
ggplot(df, aes(x = priors_count, y = as.numeric(is_recid))) +
  geom_jitter(height = 0.05, alpha = 0.3) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  theme_bw() +
  labs(title = "is_recid vs Priors Count", y = "is_recid")
```

```{r}
numericVars <- c("age", "priors_count", "is_recid")
df_numeric <- df[ , numericVars]

df_numeric$is_recid <- as.numeric(df_numeric$is_recid)

cor_matrix <- cor(df_numeric, use = "complete.obs")

print(cor_matrix)
```

We see that age and is_recid have a negative weak correlation while prior_counts and is_recid have a weak positive correlation. We can't show race because it is a categorical variable.