---
title: "Week7"
author: "Shrey Kothari"
date: "2025-11-14"
output: html_document
---

```{r}
library(tidyverse)
data <- read_csv("~/cox-violent-parsed.csv")
```
```{r}
set.seed(862331611)
rows <- nrow(data)
trainIndex <- sample(1:rows, size = 0.7*rows)
trainData <- data[trainIndex, ]
testData <- data[-trainIndex, ]
trainData <- trainData %>%
  rename(priors_count = priors_count...15)
testData <- testData %>%
  rename(priors_count = priors_count...15)
trainData <- trainData %>%
  rename(decile_score = decile_score...12)
testData <- testData %>%
  rename(decile_score = decile_score...12)
```
Step 1:
```{r}
variablesOfInterest <- c("is_recid", "race", "age", "priors_count", "juv_fel_count", "decile_score")
df <- trainData[, variablesOfInterest]
model <- lm(is_recid ~ race + juv_fel_count + age + priors_count, data = df)
summary(model)
```
Step 2:
```{r}
cor(df[, c("is_recid", "age", "priors_count", "decile_score")], use = "complete.obs")
```

```{r}
car::vif(model)
```
There are no variables with VIF that is greater than 5. We do not have redundant information.

Based on the two results above, we can keep all four variables because multicollinearity is not a concern and there are both weak positive and weak negative relationships between each variable. We could not use sex and race variables because they are categorical.

```{r}
model2 <- lm(is_recid ~ race + juv_fel_count + age + priors_count + decile_score, data = df)
summary(model2)
```
Step 3:
```{r}
plot(model2, which = 1)
```
```{r}
plot(model2, which = 2)
```
```{r}
histogram <- hist(residuals(model2), main = "Histogram of Residuals", xlab = "Residuals")
histogram
```
We see patterns of heteroscedasticity in the Residuals vs Fitted graph because we see three bands. This is because the response variable, is_recid, is a binary variable. We do see extreme outliers, such as the dot on the right side of the bottom line in the Residuals vs Fitted graph.

From the graphs, many of the assumptions have not been satisfied. Constant variance is violated because we have strong heteroscedasticity patterns. Linearity is violted because our outcome is binary. The residual are not normally distributed because we see an S shape in the QQ plot with heavy deviation from the line.

Step 4:
```{r}
pred_test <- predict(model2, newdata = testData)
RMSE <- sqrt(mean((testData$is_recid - pred_test)^2))
RMSE
```
```{r}
summary(model2)
```

Our RMSE (testing) is equal to 0.5411 while R^2 (training) is equal to 0.1324 This tells me that the model would perform poorly and has limited generalizability.

This is the sign of underfitting because we are performing poorly in both the training and testing sets.
