---
title: "Week8"
author: "Shrey Kothari"
date: "2025-11-21"
output: html_document
---

Step 1:

```{r}
library(tidyverse)
data <- read_csv("cox-violent-parsed.csv")
set.seed(862331611)
rows <- nrow(data)
trainIndex <- sample(1:rows, size = 0.7*rows)
trainData <- data[trainIndex, ]
testData <- data[-trainIndex, ]
trainData <- trainData %>%
  rename(priors_count = priors_count...15)
testData <- testData %>%
  rename(priors_count = priors_count...15)
trainData <- trainData %>%
  rename(decile_score = decile_score...12)
testData <- testData %>%
  rename(decile_score = decile_score...12)
```

```{r}
variablesOfInterest <- c("is_recid", "race", "age", "priors_count", "juv_fel_count", "decile_score")
df <- trainData[, variablesOfInterest]
model <- lm(is_recid ~ race + juv_fel_count + age + priors_count + decile_score, data = df)
summary(model)
```
Step 2:

```{r}
plot(model, which = 1)
```
```{r}
plot(model, which = 2)
```

```{r}
df$age <- df$age - mean(df$age)
df$priors_count <- df$priors_count - mean(df$priors_count)
df$juv_fel_count <- df$juv_fel_count - mean(df$juv_fel_count)

model2 <- lm(is_recid ~ race + age + I(age^2) + juv_fel_count + I(juv_fel_count^2) + priors_count +I(priors_count^2) + decile_score, data = df)
summary(model2)
```


```{r}
adj_r_squared <- summary(model2)$adj.r.squared
adj_r_squared
```
```{r}
library(car)
vif(model2)   
```
```{r}
plot(model2, which = 1)   
```

We see that in the new residuals vs fitted graph, the curvature has improved the model because the curve is nearly flat, as opposed to the line shown in the first residuals vs fitted graph.

Step 3:

One interaction I am considering between two quantitative predictors would be age and juv_fel_count. Another interaction I am considering between a quantitative and a qualitative predictor would be race and priors_count.

The interaction between age and juv_fel_count is important because the effect of having felony charges as a juvenile may vary depending on an individual’s age. Similarly, the interaction between race and priors_count allows the model to capture differences in how prior offenses influence risk of being arrested again across racial groups. Including these interactions helps the model better reflect realistic variations and improves its accuracy.

```{r}
model3 <- lm(is_recid ~ race + age + I(age^2) + juv_fel_count + I(juv_fel_count^2) + priors_count + I(priors_count^2) + decile_score + age*juv_fel_count + race*priors_count, data = df)
summary(model3)
```
Based on the model, the interaction between age and juv_fel_count is statistically significant because the p-value is equal to 0.030256. Due to the same reason with different p-values, the interaction between race and priors_count is statistically significant because each p value is below 0.05. We do see that both interactions improve fit with an increase in Adjusted R^2 and improves interpretability.

Step 4:

No, the expanded model is not too large for the sample size.

Step 5:

Our response variable Y is not strictly positive. None of the predictors require additional transformations. This is because we are already covering curvatures and non-linearity in the squared terms. It is also because we can not perform transformations on some predictors, such as decile score and race.

Step 6:

```{r}
model_final <- lm(is_recid ~ race + age + I(age^2) + juv_fel_count + I(juv_fel_count^2) + priors_count + I(priors_count^2) + decile_score + age*juv_fel_count + race*priors_count, data = df)
summary(model_final)
```
```{r}
vif(model_final)
```
```{r}
plot(model_final, which = 1)   
```


The Adjusted R^2 is equal to 0.1587, which means that the model explains 15.87 percent of the variance while accounting for the number of predictors. 

For VIF, we do see that race and the interaction between race and priors_count have values over 5, which means that we do have some multicollinearity concerns. 

In the residuals vs fitted values plot, we do see that the line has gone away from 0 again. This is most likely due to the categorical variable "race" and its interaction.

Step 7:

```{r}
pred_test <- predict(model_final, newdata = testData)
RMSE <- sqrt(mean((testData$is_recid - pred_test)^2))
RMSE
```
The Week 3 model has a lower RMSE of 0.5411 compared to the refined model’s RMSE of 0.5854 on the test data. This means the Week 3 model actually predicts more accurately.

